Attrition and Achievement Gaps in Online Learning

### 2015-01-24 paper_v2.Rnw
updated version

\title{Attrition in Massive Open Online Courses}

@@

\begin{abstract}
Attrition in online learning is generally higher than in traditional settings. High attrition in large-scale open learning environments prompted a shift from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation.
A systematic analysis of individual differences in attrition in 17 diverse massive open online courses ($N>60,000$) revealed substantial achievement gaps: women were less likely to persist than men, and learners in Africa, Asia, and South America persisted less and scored lower than learners in Europe, North America, and Oceania. More persistent learners were, however, only marginally more satisfied with their achievement. 
Successful online learners who left the course early reported higher levels of goal striving, feelings of social belonging, and growth mindset than unsuccessful ones ($N=756$). The shortage of time was the primary obstacle for most learners, and many were hindered by insufficient volitional control.
A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
\end{abstract}

\keywords{Online learning, MOOC, persistence, dropout, individual differences, volition, social belonging, mindset}

@@

This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses overall. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who reported having not enough time, and examined goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

\begin{table*}[th]
\caption{Summary statistics of learner demographics and behavior in 17 MOOCs}
\small
\center
\label{tab:s1sum}
\begin{tabular}{llccccccccccc}
% & & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
% & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
\toprule
 &            &          & \multicolumn{3}{c}{Survey Responses} & & & & & \multicolumn{3}{c}{Videos Watched} \\
 \cmidrule(r){4-6} \cmidrule(r){11-13}
 & Topic Area & Enrolled & Initial & Feedback & Both & Female & Age$^1$   & Education$^2$ & Dropout$^3$ & $>$ 30\% & $>$ 50\% & $>$ 80\% \\
\midrule
C1 & Business & 61,096   & 9,943  & 2,367    & 1,282    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
C2 & Business & 41,081   & 6,544  & 883      & 630      & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
C3 & CS       & 27,104   & 6,104 & 2,854     & 2,116    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
C4 & CS       & 56,154   & 3,406 & 1,123     & 352      & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
C5 & CS       & 57,636   & 5738  & 402       & 342      & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
C6 & Engineering & 16,220 & 2,540 & 269      & 182      & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
C7 & Enginerring & 7,178 & 1,312 & 202       & 167      & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
C8 & Humanities & 7,634 & 3,631  & 152       & 127      & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
C9 & Humanities & 6,475 & 3,315  & 326       & 262      & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
C10 & Humanities & 9,903 & 4,227 & 1,570     & 1,066    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
C11 & Humanities & 630 & 167     & 80        & 40       & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
C12 & Math \& Stats & 49,317 & 3,090 & 862   & 510      & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 4,096   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
C14 & Math \& Stats & 16,857 & 2,892 & 363   & 282      & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 1,521   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
C16 & Math \& Stats & 27,236 & 2,711 & 637   & 610      & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
C17 & Math \& Stats & 13,132 & 3,629 & 843   & 626      & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
\bottomrule
\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelor's or higher degree  $^3$ self-reported in the feedback survey}
\end{tabular}
\end{table*}

@@

Research and theorizing on attrition in education has a rich history. This review is intended to serve as a foundation to build on with the current research. We therefore simultaneously develop research questions and hypotheses to investigate and test in the current work. The focus of this brief review is on how ways of thinking about attrition have evolved over the last decades.

@@

\begin{description}
\item[RQ3a] What reasons do learners report for disengaging from a course?
\item[RQ3b] How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
\end{description}

@@

The present study extends the existing literature on persistence with a representative quantitative treatment of attrition in MOOCs. We address the research questions developed in the previous section with a combination of self-report and behavioral data from 17 diverse MOOCs. Going beyond specific research questions, this observational analysis is intended to inform our understanding of attrition in MOOCs to better support learners in these environments. This can be achieved through general changes, for instance, to the course format; or by targeted interventions that support a particular group of learners, such as those who exhibit low levels of goal striving (cf. Study 2).

@@

Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (cf. Table \ref{tab:s1sum}). For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment. 
% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
On the initial course surveys, learners reported their gender, age, education, prior experience with the topic and MOOCs, enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}), and how many hours per week they intend to spend on the course. On the feedback survey, learners reported their satisfaction with the course (6-item index, range from 1 to 7, $M=5.29, SD=0.94, \alpha=.82$), and whether they `stopped participating in the course before it ended' or `remained active until the end'. Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (Figure \ref{fig:s1reason} lists reasons).\footnote{Instead of a select-all-that-apply question, learners were encouraged to consider each reason in turn by using binary-choice items. The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to mitigate acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region.

Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course. Behavioral indicators were chosen to represent course milestones: watching over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile. A regression analysis predicting course milestones and performance was performed based on a subset of learners who completed the initial course survey in the 17 MOOCs in our dataset. Out of the 63,620 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$. Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.\footnote{Multiple imputation and subsequent pooled analyses were performed with the $R$ mice package.}
% The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
% The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
% Information about age, gender, and education level were only available for learners who completed the initial course survey.
% We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior

\subsection{Results}

We address the research questions developed in the previous section in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across subsets of learners ($RQ2ab$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3ab$).

\subsubsection{Individual Differences in Attrition}

We fit a logistic mixed-effects (hierarchical) model to predict multiple binary persistence and performance metrics (milestones) to estimate individual differences in learner characteristics, including demographic and geographic features. Correlations between predictors were low ($|r|<0.16$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded. The coefficient on education, for instance, reflects the association between having a bachelor's degree and achieving a milestone, partialing out all other predictors, including age. Fourteen MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.\footnote{Data for 3 of the 17 MOOCs was unavailable due to anonymization.}

Figure \ref{fig:s1coefs} presents transformed regression coefficients with 95\% confidence intervals from each model. The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population. The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average aged male learners in North America without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it. In this sample of survey respondents, the baseline probability for each milestone was high: videos watched (27\% above 30\%, 16\% above 50\%, 8\% above 80\%) assessments attempted (78\% above 30\%, 55\% above 50\%, 33\% above 80\%), and grade (42\% above the 60$^\text{th}$, 31\% above the 80$^\text{th}$ percentile).

A number of significant and substantial individual differences emerged. Women were 10 to 15\% less likely than men to persist with lectures and assessments. Women, who constituted 41\% of learners in the sample, were also less likely to score a grade above the 60$^\text{th}$ percentile than men. Learners in Africa ($n = 2,097$), Asia ($n = 8,199$), and South America ($n = 2,574$) were 20 to 50\% less likely to persist in video watching and tended to achieve lower grades than learners in North America ($n = 18,353$). In contrast, learners in Europe ($n = 11,992$) exhibited 20 to 50\% higher persistence with lectures and assessemnts than learners in North America. While learners in Asia exhibited lower persistence with videos, they were more likely to attempt more assessments than learners from North America. Learners in Oceania ($n = 1,166$) performed similarly to those in North America and Europe.

Besides gender and geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance. More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to score highly on them. Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance. Learners who took prior MOOCs, however, were less likely to persist, yet achieved higher grades.

<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Estimates from logistic mixed-effects model predicting persistence and performance milestones. An estimate of -10 indicates a 10\\% lower chance of achieving a milestone.'>>=
library(ggplot2)
res = read.csv("pred_coefs.csv")

res$y = factor(res$y, levels=c(">30% videos", ">50% videos", ">80% videos", ">60% pctile grade", ">30% assessments", ">50% assessments", ">80% assessments", ">80% pctile grade"))
res$y = factor(res$y, labels=c("watched >30% videos", "watched >50% videos", "watched >80% videos", ">60th percentile grade", "attempted >30% assessments", "attempted >50% assessments", "attempted >80% assessments", ">80th percentile grade"))

res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:S.America")))
res$coef2 = factor(res$coef2, labels=rev(c("Intercept", "Gender: Female", "Age (normalized)", "Education: Bachelor+", "Prior Experience", "Intent to complete", "Hours: 2-4", "Hours: 4+ ", "Prior MOOCs: 1-3", "Prior MOOCs: 4+ ", "Location: Europe", "Location: Oceania", "Location: Africa", "Location: Asia", "Location: South America")))

@@

# This is what it should be
ggplot(subset(res, coef2!="Intercept"), 
  aes(coef2, 100*(exp(est)-1), ymin=100*(exp(est-1.96*se)-1), ymax=100*(exp(est+1.96*se)-1), color=factor(sign))) + 
  geom_hline(yintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
  theme_minimal() + theme(legend.position="none",
                          axis.title=element_text(vjust=0),
                          panel.border=element_rect(fill="transparent",size=.5,color="grey50")) + 
  scale_color_manual(values=c('black','grey50','black')) +
  scale_y_continuous(breaks=c(-60,-40,-20,0,20,40,60,80)) +
  labs(y="Predicted change in probability of achieving each `milestone` (95% CIs)", x="")
coefs = subset(res, coef2=="Intercept", select=c("est","y"))  # should report these in the text of reference
coefs$exp_est = round(100*exp(coefs$est)) # estimates seem high
# >30% videos      27
# >50% videos      16
# >80% videos       8
# >30% assessments      78
# >50% assessments      55
# >80% assessments      33
# >60% pctile grade      42
# >80% pctile grade      31
@

\subsubsection{Learner Satisfaction}
% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?

Attrition can be a sign of success if a learner has achieved their personal goal. We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who participanted to different degrees ($RQ2b$). Satisfaction ratings were compared with a mixed-effects regression to model differences across courses and generalize to a larger population of MOOCs. Learners who reported that they stopped participating were 9\% less satisfied than those who reported not to have stopped ($t_{15140}=18, p<.001$). Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied. While the the proportion of videos significantly predicted higher satisfaction ($t_{15700}=14, p<.001$), the effect was relatively weak. Those who watched over 75\% of videos were only 5\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=4.97, M_{>\text{75\%}}=5.22$).

<<study1_satis_si_score_dist_comparison, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=5, fig.cap='Satisfaction index scores for learners who reported they stopped / did not stop participating before end of course'>>=
#b = a[!is.na(a$did_dropout),]
#ggplot(b) + geom_boxplot(aes(x=b$did_dropout,y=b$satisfaction)) + theme_bw() + labs(y="Satisfaction",x="Stopped participating?")
@

% We also examined differences in satisfaction between learners grouped by certain behaviors. First, we classified learners into 4 groups based on the fraction of course videos they viewed. The box plots for the satisfaction scores in each group are shown in figure \ref{fig:study1_satsfaction_by_vid_groups}. The mean satisfaction score increases slightly as the number of videos viewed by the learners increases, with the exception of the scores reported by learners in the 0-25\% group. Possible reasons for this are learners who join only to view a small part of the course or learners who join merely to explore.

<<echo=FALSE,eval=FALSE>>=
library(dplyr)
sat = read.csv("satisfaction.csv")
sat$num_videos_watched[is.na(sat$num_videos_watched)] = 0
sat = sat %>% group_by(course_id) %>% 
  mutate(prop_vid = num_videos_watched / quantile(num_videos_watched[num_videos_watched!=0], .90))

sat$prop_vid[sat$prop_vid>1] = 1
sat$prop_vid4 = cut(sat$prop_vid, breaks=c(-Inf, .25, .5, .75, Inf))
table(sat$prop_vid4)

summary(lmer(satisfaction~ did_dropout + (1|course_id), sat))
5.170/(5.170-0.4456)

summary(lmer(satisfaction~prop_vid4 + (1|course_id), sat))
summary(lmer(satisfaction~prop_vid + (1|course_id), sat))
(4.969+.2466)/4.969
@

@@

% \subsubsection{Predicting Likely Dropouts}

% SHERIF TODO: Add results of prediction. How well did the prediction work? And who took the survey; how sure was the model that those who took the survey would actually drop out, relative to those who didn't take the survey?

@@

@@

\subsection{Discussion}

>>>>

\title{Attrition and Achievement Gaps in Online Learning}

@@

\begin{abstract}
Attrition in online learning is generally higher than in traditional settings, especially in large-scale online learning environments. This has prompted a shift from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation.
A systematic analysis of individual differences in attrition in 20 diverse massive open online courses ($N>67,000$) revealed a global achievement gap and a gender achievement gap. Women and learners in Africa, Asia, and Latin America scored lower grades and persisted less than men and learners in Europe, Oceania, and Northern America. More persistent learners were, however, only marginally more satisfied with their achievement. 
Successful online learners who left the course early reported higher levels of goal striving, feelings of social belonging, and growth mindset than unsuccessful ones ($N=756$). The shortage of time was the primary obstacle for most learners, and many were hindered by insufficient volitional control.
A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
\end{abstract}

\keywords{Massive open online course, MOOC, persistence, dropout, individual differences, volition, social belonging, mindset}

@@

This paper presents a systematic investigation of attrition in online learning based on self-report and behavioral data collected from over 100,000 learners in 21 courses in total. Identifying characteristics associated with attrition and reasons for disengaging can facilitate earlier recognition of learners struggeling to achieve their learning goals. These learners could benefit from changes to the course or targeted interventions. We briefly review the large literature on attrition in educational environments, with a focus on key developments in understanding the causes of attrition. Building on this review, Study 1 offers insights into reasons for disengaging from MOOCs, learners' satisfaction, and individual differences in attrition and performance. For Study 2, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. We examined differences in psychological factors between subjectively successful and unsuccessful learners, and investigated the role of volitional control for those who reported having not enough time. Learners' open responses about challenges they faced while taking the course were iteratively coded to develop an empirically-grounded list of reasons for attrition.

\begin{table*}[th]
\caption{Summary statistics for 20 massive open online courses}
\small
\center
\label{tab:s1sum}
\begin{tabular}{llccccccccccc}
% & & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
% & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
\toprule
 &            &          & \multicolumn{3}{c}{Survey Responses} & & & & & \multicolumn{3}{c}{Videos Watched} \\
 \cmidrule(r){4-6} \cmidrule(r){11-13}
 & Topic Area & Enrolled & Initial & Feedback & Both & Female & Age$^1$   & Education$^2$ & Dropout$^3$ & $>$ 30\% & $>$ 50\% & $>$ 80\% \\
\midrule
C1 & Business & 61,233 & 8,618 & 1,796 & 1,282 & 47\% & 36 (28,43) & 94\% & 11\% & 11\% & 9\% & 6\%\\
C2 & Business & 45,202 & 5,951 & 789 & 631 & 43\% & 36 (27,42) & 90\% & 10\% & 11\% & 8\% & 6\%\\
C3 & Computer Science & 28,086 & 5,918 & 2,815 & 2,295 & 36\% & 33 (23,40) & 63\% & 3\% & 15\% & 11\% & 8\%\\
C4 & Computer Science & 56,314 & 2,686 & 749 & 352 & 14\% & 33 (25,39) & 79\% & 21\% & 13\% & 10\% & 8\%\\
C5 & Computer Science & 79,273 & 7,319 & 1,740 & 1,236 & 16\% & 31 (24,36) & 81\% & 11\% & 15\% & 11\% & 8\%\\
C6 & Education & 40,214 & 24,192 & 5,174 & 4,969 & 56\% & 28 (15,40) & n.a. & 5\% & 17\% & 13\% & 8\%\\
C7 & Education & 5,325 & 2,429 & 624 & 615 & 80\% & 43 (35,50) & 99\% & 2\% & 24\% & 17\% & 12\%\\
C8 & Engineering & 11,958 & 2,341 & 262 & 196 & 16\% & 35 (26,43) & 84\% & 16\% & 14\% & 9\% & 6\%\\
C9 & Engineering & 5,701 & 5,652 & 396 & 347 & 16\% & 33 (24,38) & 73\% & 16\% & 18\% & 11\% & 7\%\\
C10 & Engineering & 7,139 & 1,250 & 197 & 168 & 17\% & 31 (22,36) & 72\% & 7\% & 11\% & 8\% & 6\%\\
C11 & Humanities & 7,656 & 2,512 & 142 & 131 & 89\% & 33 (24,38) & 82\% & 9\% & 14\% & 9\% & 6\%\\
C12 & Humanities & 4,949 & 2,921 & 292 & 270 & 91\% & 33 (25,38) & 83\% & 17\% & 15\% & 10\% & 7\%\\
C13 & Mathematics & 49,509 & 3,031 & 836 & 510 & 35\% & 37 (26,47) & 76\% & 32\% & 12\% & 9\% & 6\%\\
C14 & Mathematics & 13,310 & 2,861 & 346 & 298 & 10\% & 31 (25,34) & 93\% & 15\% & 15\% & 10\% & 6\%\\
C15 & Natural Sciences & 9,918 & 3,957 & 1,351 & 1,223 & 51\% & 37 (26,47) & 78\% & 4\% & 19\% & 15\% & 11\%\\
C16 & Natural Sciences & 543 & 163 & 79 & 50 & 47\% & 34 (26,39) & 98\% & 18\% & 27\% & 20\% & 12\%\\
C17 & Physics & 6,452 & 1,069 & 269 & 269 & 14\% & 36 (25,46) & 82\% & 11\% & 15\% & 10\% & 6\%\\
C18 & Statistics & 11,733 & 11,465 & 1,837 & 1,569 & 22\% & 36 (28,41) & 96\% & 9\% & 34\% & 25\% & 15\%\\
C19 & Statistics & 13,161 & 3,495 & 819 & 627 & n.a. & n.a. & n.a. & 6\% & 13\% & 9\% & 7\%\\
C20 & Writing & 35,374 & 12,514 & 2,877 & 2,876 & n.a. & n.a. & n.a. & 8\% & 23\% & 14\% & 8\%\\
\bottomrule
\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelor's or higher degree  $^3$ self-reported in the feedback survey}
\end{tabular}
\end{table*}

@@

Research and theory on attrition has a rich history in education. This review focuses on how conceptions of attrition have evolved over the last decades.  As this review is intended to serve as a foundation to build on with the current research, we concurrently develop hypotheses and research questions to investigate.

@@

\begin{description}
\item[RQ3a] What reasons do learners report for disengaging from a course?
\item[RQ3b] How do reasons for disengaging vary by learner characteristics?
\end{description}

@@

The present study extends the existing literature on attrition with a large-scale quantitative treatment of attrition in MOOCs. We address the research questions developed in the previous section with the larger objective of informing our understanding of attrition in online courses to better support learners. This could be achieved through general changes to the course content or presentation; or, using targeted interventions that support particular groups of struggeling learners.

@@

Twenty MOOCs on topics in a variety of disciplines were selected for investigation. Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course. For each course, we combined data from an initial course survey (distributed in the first weeks of the course), a feedback survey (distributed in the last weeks of the course), and behavioral data collected in the learning environment. On the initial course survey, learners reported their gender, age, education, prior experience with the topic and MOOCs, and how many hours per week they intend to spend on the course. On the feedback survey, learners reported their satisfaction with the course (5-item index, possible range from 0 to 5.2, $M=3.95, SD=0.81, \alpha=.81$), and whether they `stopped participating in the course before it ended' or `remained active until the end'. Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (cf. Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, this question style encouraged learners to consider each reason separately. The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to reduce acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region. Behavioral indicators were chosen to represent course milestones: watching (or downloading) over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.

To analyze individual differences, we fit a logistic mixed-effects (hierarchical) model predicting binary persistence and performance milestones using learner characteristics, including demographic and geographic features. The analysis included 67,333 learners who completed the initial course survey, but excluded 381 learners who left all questions unanswered and learners in courses C6, C7, C19, C20 due to missing survey items. Any remaining missing survey responses were multiply imputed using other survey responses. We report pooled estimates that account for added uncertainty.\footnote{Multiple imputation ($m=10$) using the $R$ mice package.} Correlations between predictors were low ($|r|<0.15$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded. Courses were modeled as a random effect (instead of including dummy variables) to generalize to a larger population of similar MOOCs.

\subsection{Results and Discussion}

We address the posed research questions in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across learner subgroups ($RQ2$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3$).

\subsubsection{Individual Differences in Attrition and Performance}

Individual differences in attrition and performance are illustrated in Figure \ref{fig:s1coefs} as transformed regression coefficients with 95\% confidence intervals for each milestone. The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population. The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average-aged male learners in Northern America\footnote{Learners in the United States, Canada, Bermuda, Greenland.} without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it. The baseline probability for each milestone was relatively high: lectures watched (20\% above 30\%, 11\% above 50\%, 4\% above 80\%), assessments attempted (37\% above 30\%, 25\% above 50\%, 18\% above 80\%), and grades (30\% above the 60$^\text{th}$, 22\% above the 80$^\text{th}$ percentile).

A number of significant and substantial individual differences emerged. Women were 12 to 20\% less likely than men to persist with lectures and assessments. Women, who constituted 34\% of learners in the sample, were also 10\% (7\%) less likely than men to score a grade above the 60$^\text{th}$ (80$^\text{th}$) percentile.  Moreover, learners in Africa ($N = 2,458$), Asia ($N = 13,280$), and Latin America ($N = 5,344$) were 24 to 50\% less likely to persist with lectures and 31 to 52\% less likely to achieve grade milestones than learners in Northern America ($N = 26,651$). In contrast, learners in Europe ($N = 17,988$) exhibited 16 to 30\% higher persistence with lectures and assessemnts than learners in Northern America. While learners in Asia exhibited lower persistence with videos, they were at least as likely to attempt assessments as learners in North America. Learners in Oceania ($N = 1,612$) performed similarly to those in Northern America and Europe. Besides gender and geography, older learner were more likely to persist with lectures and assessments, but achieved lower grades than younger learners. More educated learners, those with a bachelor's or higher degree, were 15 to 25\% more likely to persist and 10\% more likely to score above the 60$^\text{th}$ percentile. Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance. Learners who participated in prior MOOCs had mixed persistence patterns, but achieved higher grades.

The large scale of MOOCs has revealed a global achievement gap and a gender achievement gap in online learning. Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America. There was a good representation from each geographic region in the dataset and the courses under consideration were on diverse topics, including some in the humanities and business with high female participation rates. The strength and significance of these relationships is remarkable, in particular with several additional demographic, geographic, and preference measures in the model. The first generation of large-scale online learning, characterized by MOOCs, is apparently not immune to inequalities that have pervaded traditional educational settings. The present work cannot move beyond presenting evidence for the existence of these gaps and discussing potential causes. Of course there exists no innate feature of women or people in certain locations that is responsible for these gaps; but rather, gender and location are associated with countless individual characteristics, some of which impact academic achievement. The achievement gaps could plausibly result from differences in Internet access, language barriers, or from feelings of psychological threat, such as fears of confirming a negative stereotype \cite{spencer1999stereotype} or not belonging in the course \cite{walton2007question}. More work is needed to identify potential causal mechanisms underlying these gaps.

<<s1coefs, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=6, fig.width=10, fig.cap='Individual differnces in persistence and course grades showing substantial achievement gaps. An estimate of -10 indicates a 10\\% lower chance of achieving a milestone.'>>=
library(ggplot2)
res = read.csv("pred_coefs2014-1-15.csv")

res$y = factor(res$y, levels=c(">30% videos", ">50% videos", ">80% videos", ">60% pctile grade", ">30% assessments", ">50% assessments", ">80% assessments", ">80% pctile grade"))
res$y = factor(res$y, labels=c("> 30% lectures", "> 50% lectures", "> 80% lectures", "> 60th percentile grade", "> 30% assessments", "> 50% assessments", "> 80% assessments", "> 80th percentile grade"))

res$coef2 = factor(res$coef2, levels=rev(c("Intercept", "Gender:F", "Age (norm.)", "Educ.:Bachelor+", "Prior Exp.", "Intent:All", "Hours:2-4", "Hours:4+ ", "Prior MOOCs:1-3", "Prior MOOCs:4+ ", "Geo:Europe", "Geo:Oceania", "Geo:Africa", "Geo:Asia", "Geo:Latin America")))
res$coef2 = factor(res$coef2, labels=rev(c("Intercept", "Gender: Female", "Age (normalized)", "Education: Bachelor+", "Prior Experience", "Intent to complete", "Hours: 2-4", "Hours: 4+ ", "Prior MOOCs: 1-3", "Prior MOOCs: 4+ ", "Location: Europe", "Location: Oceania", "Location: Africa", "Location: Asia", "Location: Latin America")))

@@

# This is what it should be
ggplot(subset(res, coef2!="Intercept"), 
  aes(coef2, 100*(exp(est)-1), ymin=100*(exp(est-1.96*se)-1), ymax=100*(exp(est+1.96*se)-1), color=factor(sign))) + 
  geom_hline(yintercept=0) + geom_pointrange() + facet_wrap(~y, ncol=4) + coord_flip() + 
  theme_minimal() + theme(legend.position="none",
                          axis.title=element_text(vjust=0),
                          panel.border=element_rect(fill="transparent",size=.5,color="grey50")) + 
  scale_color_manual(values=c('black','grey50','black')) +
  scale_y_continuous(breaks=c(-60,-40,-20,0,20,40,60,80)) +
  labs(y="Predicted change in probability of achieving each milestone (95% CIs)", x="")
coefs = subset(res, coef2=="Intercept", select=c("est","y"))  # should report these in the text of reference
coefs$exp_est = round(100*exp(coefs$est)) # estimates seem high
# 1   -1.63          > 30% lectures      20
# 16  -2.20          > 50% lectures      11
# 31  -3.13          > 80% lectures       4
# 46  -1.00       > 30% assessments      37
# 61  -1.38       > 50% assessments      25
# 76  -1.73       > 80% assessments      18
# 91  -1.21 > 60th percentile grade      30
# 106 -1.51 > 80th percentile grade      22
@

\subsubsection{Satisfaction}
% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?

Attrition can be a sign of success if a learner has achieved their personal learning objectives. We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who persisted to different extents ($RQ2b$). Satisfaction ratings were compared using mixed-effects regression ($N=23,390$). Learners who reported that they stopped participating were 12\% less satisfied than those who reported not to have stopped ($t_{453}=21, p<.001$). Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied. The proportion of videos significantly predicted higher satisfaction ($t_{101}=20, p<.001$), such that those who watched over 75\% of videos were 8\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=3.80, M_{>\text{75\%}}=4.09$). As levels of self-reported satisfaction were generally high, it remains unclear if the estimated differences in satisfaction are consequential. Learners who disengaged early were only somewhat less satisfied than those who persisted in the course, which suggests that some learners arrived with modest course goals and achieved them. The following analysis of reasons for attrition can shed more light on this hypothesis.

@@

@@

Personal success in a MOOC was associated with three established psychological factors: goal striving, social belonging, and growth mindset. Prior work in more traditional achievement-oriented environments found reduced levels of these psychological factors to result in lower persistence and performance. The current observational findings highlight the critical role of these factors in an online learning environment. Learners with higher levels of goal striving, social belonging, and growth mindset reported greater personal success ($H1,H2,H3$). An online learner with, for instance, a strong sense of social belonging may derive greater satisfaction from the course and persist for longer than one with a weak sense of belonging. These psychological factors tend to vary across poeple based on prior experiences and they can change as a result of contextual cues. Pschological factors could thereby, at least partially, account for the individual differences in persistence and performance observed in Study 1.

@@
